Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-18 manuscript/draft.Rmd
Editorial work on Intro
Includes moving Table 1 into methods where the detail present in it makes more
sense. Replaces the reference to that table in the intro with a reference to
Figure 1, which works better there.

Forecasting the future state of ecological systems is increasingly important for
planning and management, and also for quantitatively evaluating how well our
models capture the key ecological processes governing these systems [@clark2001;
@houlahan2017; @dietze2017]. Forecasts regarding biodiversity are especially
important, due to biodiversity's central role in conservation planning and its
sensitivity to anthropogenic effects [@cardinale2012; @diaz2015; @tilman2017].
High-profile studies forecasting large biodiversity declines over the coming
decades have played a large role in shaping ecologists' priorities [as well as
those of policymakers; e.g. @ipcc2014], but it is inherently difficult to
evaluate such long-term predictions before the projected biodiversity declines
have occurred.

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
(SDMs). In general, these models describe individual species' occurrence
patterns as functions of the environment [@thomas2004; @urban2015]. By plugging
forecasts of future environmental conditions into these functions, ecologists
can make predictions about where each species will occur in the future. These
species-level predictions can then be added up ("stacked") to generate predicted
richness values [e.g. @calabrese2014]. Alternatively, models that directly
relate spatial patterns of species richness to the environment have been
developed and generally perform equivalently to stacked SDMs [@algar2009;
@distler2015]. This approach is sometimes referred to as "macroecological"
modeling, because it models the larger-scale pattern (richness) directly
[@distler2015].

Despite the emerging interest in forecasting species richness and other aspects
of biodiversity [[refs]], little is known about how effectively we can
anticipate these dynamics. This is due in part to the long time scales over
which many ecological forecasts are applied [and the resulting difficulty in
assessing whether the predicted changes occurred; @dietze2016]. What we do know
comes from a small number of hindcasting studies, where models are built using
data on species occurrence and richness from the past and evaluated on their
ability to predict contemporary patterns [e.g., @algar2009; @distler2015]. These
studies are a valuable first step, but lack several components that are
important for developing forecasting models with high predictive accuracy, and
for understanding how well different methods can predict the future. The
components an effective forecasting and evaluation strategy (Box 1) broadly
involve: 1) expanding the use of data to include biological and environmental
time-series [@tredennick2016]; 2) accounting for uncertainty in observations,
processes, model choice, and forecast evaluation [@clark2001; @dietze2017]; and
3) conducting meaningful evaluations of the forecasts by hindcasting, archiving
short-term forecasts, and comparing forecasts to baselines to determine whether
the forecasts are more accurate than simply assuming the system is basically
static [@perretti2013].

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,000 of sites located throughout North America, while following the best
practices in Box 1 for ecological forecasting. To do this, we combine 32 years
of time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. Datasets that
span a time scale of 30 years or more have only recently become available for
large-scale time-series based forecasting. A dataset of this size allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and two simple baselines that predict
constant richness for each site, on average (Table 1). All of our forecasting
models account for uncertainty and observation error, are evaluated across
different time lags using hindcasting, and are publicly archived to allow future
assessment. We discuss the implications of these practices for our understanding
of, and confidence in, the resulting forecasts, and how we can continue to build
on these approaches to improve ecological forecasting in the future.

@@

**Multi-timescale forecasts.** Another challenge for ecological forecasting is
the timescales of typical forecasts. We evaluated forecasts using yearly
timesteps up to a decade into the future. In contrast, many ecological forecasts
are temporally aggregated to 5-32 year timesteps and projected up to a century
into the future. These are commonly beyond the career or even lifespan of the
researchers, which makes it impossible to directly validate the models'
predictions. Currently, assessments at these long time-scales can only be made
with a small number of opportune datasets. For example, hindcasts have been used
to assess models of species richness [@algar2009; @distler2015] and distribution
[@rapacciuolo2012; @moran-ordonez2017; @araujo2005; @eskildsen2013] using data
aggregated across similar time scales to those studied here. More evaluation
studies like these are needed but there is a paucity of adequate long term data
available. Short term forecasts will allow for relatively rapid validation of
models by researchers without needing to wait for enough data to accumulate
[@tredennick2016; @dietze2016]. This does not mean that they should replace
large scale long-term forecasts. As discussed above, drivers of species richness
are likely to differ at different temporal scales [@rosenzweig1995; @white2004;
@white2007]. Thus, short-term forecasts should be expected to add value to
scientists and policymakers, and inform (not replace) long-term forecasts.
Hopefully, by evaluating trends in the shorter term forecasts as done here it
will be possible to also use the information from short-term forecasting to
improve longer-term estimates.

>>>>

Forecasting the future state of ecological systems is increasingly important for
planning and management, and also for quantitatively evaluating how well
ecological models capture the key processes governing natural systems
[@clark2001; @houlahan2017; @dietze2017]. Forecasts regarding biodiversity are
especially important, due to biodiversity's central role in conservation
planning and its sensitivity to anthropogenic effects [@cardinale2012;
@diaz2015; @tilman2017]. High-profile studies forecasting large biodiversity
declines over the coming decades have played a large role in shaping ecologists'
priorities [as well as those of policymakers; e.g. @ipcc2014], but it is
inherently difficult to evaluate such long-term predictions before the projected
biodiversity declines have occurred.

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
[SDMs; @thomas2004; @urban2015]. In general, these models describe individual
species' occurrence patterns as functions of the environment. Predictions about
where each species will occur in the future are made by using forecasts of
future environmental conditions to generate predictions from these models. These
species-level predictions are then combined ("stacked") to generate forecasts
for species richness [e.g. @calabrese2014]. Alternatively, models that directly
relate spatial patterns of species richness to environment conditions have been
developed and generally perform equivalently to stacked SDMs [@algar2009;
@distler2015]. This approach is sometimes referred to as "macroecological"
modeling, because it models the larger-scale pattern (richness) directly
[@distler2015].

Despite the emerging interest in forecasting species richness and other aspects
of biodiversity [[refs]], little is known about how effectively we can
anticipate these dynamics. This is due in part to the long time scales over
which many ecological forecasts are applied [and the resulting difficulty in
assessing whether the predicted changes occurred; @dietze2016]. What we do know
comes from a small number of hindcasting studies, where models are built using
data on species occurrence and richness from the past and evaluated on their
ability to predict contemporary patterns [e.g., @algar2009; @distler2015]. These
studies are a valuable first step, but lack several components that are
important for developing forecasting models with high predictive accuracy, and
for understanding how well different methods can predict the future. These "best
practices" for effective forecasting and evaluation (Box 1) broadly involve: 1)
expanding the use of data to include biological and environmental time-series
[@tredennick2016]; 2) accounting for uncertainty in observations, processes,
model choice, and forecast evaluation [@clark2001; @dietze2017]; and 3)
conducting meaningful evaluations of the forecasts by hindcasting, archiving
short-term forecasts, and comparing forecasts to baselines to determine whether
the forecasts are more accurate than assuming the system is basically static
[@perretti2013].

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,200 of sites located throughout North America, while following best
practices for ecological forecasting (Box 1). To do this, we combine 32 years of
time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. Datasets that
span a time scale of 30 years or more have only recently become available for
large-scale time-series based forecasting. A dataset of this size allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and two simple baselines that predict
constant richness for each site, on average (Figure 1). All of our forecasting
models account for uncertainty and observation error, are evaluated across
different time lags using hindcasting, and are publicly archived to allow future
assessment. We discuss the implications of these practices for our understanding
of, and confidence in, the resulting forecasts, and how we can continue to build
on these approaches to improve ecological forecasting in the future.

@@

**Multi-timescale forecasts.** Another challenge for ecological forecasting is
the timescales of typical forecasts. We evaluated forecasts using yearly
timesteps up to a decade into the future. In contrast, many ecological forecasts
are temporally aggregated to 5-30 year timesteps and projected up to a century
into the future. These are commonly beyond the career or even lifespan of the
researchers, which makes it impossible to directly validate the models'
predictions. Currently, assessments at these long time-scales can only be made
with a small number of opportune datasets. For example, hindcasts have been used
to assess models of species richness [@algar2009; @distler2015] and distribution
[@rapacciuolo2012; @moran-ordonez2017; @araujo2005; @eskildsen2013] using data
aggregated across similar time scales to those studied here. More evaluation
studies like these are needed but there is a paucity of adequate long term data
available. Short term forecasts will allow for relatively rapid validation of
models by researchers without needing to wait for enough data to accumulate
[@tredennick2016; @dietze2016]. This does not mean that they should replace
large scale long-term forecasts. As discussed above, drivers of species richness
are likely to differ at different temporal scales [@rosenzweig1995; @white2004;
@white2007]. Thus, short-term forecasts should be expected to add value to
scientists and policymakers, and inform (not replace) long-term forecasts.
Hopefully, by evaluating trends in the shorter term forecasts as done here it
will be possible to also use the information from short-term forecasting to
improve longer-term estimates.
